---
title: "`r params$report_title`"
output: 
  html_document:
    toc: true
    toc_float: false
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: none
params:
  report_title: "BCB-SR RNA-seq Report"  # default fallback
  params_file: 
pandoc_args: 
  - "+RTS" 
  - "-K16000m"
  - "-RTS"
date: "`r format(Sys.time(), '%m/%d/%y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.align="center")
library(ggplot2)
library(gridExtra)
library(DT)
library(limma)
library(edgeR)
library(qs)
library(kableExtra)
library(yaml)
source("R/HRK_funcs.R")
library(downloadthis)
select <- dplyr::select
#TEST=TRUE
```

```{r params}
parse_params <- function(filepath) {
  lines <- readLines(filepath)
  param_list <- list()
  
  for (line in lines) {
    line <- trimws(line)
    
    # Skip empty lines and comments
    if (line == "" || startsWith(line, "#")) {
      next
    }
    
    if (grepl("^--", line)) {
      space_pos <- regexpr("\\s", line)
      
      if (space_pos > 0) {
        key <- gsub("^--", "", substr(line, 1, space_pos - 1))
        
        # NORMALIZE HYPHENS TO UNDERSCORES for consistency
        key <- gsub("-", "_", key)
        
        value <- trimws(substr(line, space_pos + 1, nchar(line)))
        
        # Remove surrounding quotes
        if ((startsWith(value, '"') && endsWith(value, '"')) ||
            (startsWith(value, "'") && endsWith(value, "'"))) {
          value <- substr(value, 2, nchar(value) - 1)
        }
        
        # Convert logical strings
        if (tolower(value) == "true") {
          value <- TRUE
        } else if (tolower(value) == "false") {
          value <- FALSE
        }
        
        param_list[[key]] <- value
      }
    }
  }
  
  return(param_list)
}

# Determine which params file to use
if (!exists("TEST") || is.null(TEST)) {
  # Production mode: read from passed-in params
  if (!is.null(params$params_file) && file.exists(params$params_file)) {
    report_params <- parse_params(params$params_file)
    
    # CRITICAL: Merge with directly-passed params, giving them priority
    # This handles report_title and any other params passed via sbatch
    for (param_name in names(params)) {
      if (param_name != "params_file") {  # Don't overwrite the file path itself
        report_params[[param_name]] <- params[[param_name]]
      }
    }
    
  } else {
    stop("No valid params file provided via params$params_file")
  }
} else {
  # Testing mode
  report_params <- parse_params("/blue/cancercenter-dept/hkates/Tests/rnareport/output/app-test-3_params.txt")
}

# DEBUG: Print key params to verify loading
cat("DEBUG: annotation_db =", report_params$annotation_db, "\n")
cat("DEBUG: report_title =", report_params$report_title, "\n")
cat("DEBUG: output_path =", ifelse(!is.null(report_params$output_path), 
                                    report_params$output_path, 
                                    "NOT SET"), "\n")
```

```{r analysis}
# Source the function definition
source("R/analysis.R")

# Run the analysis and extract results
analysis_results <- run_rnaseq_analysis(report_params)

# Unpack results into global environment for backwards compatibility
list2env(analysis_results, envir = .GlobalEnv)
```

# Project Summary

**PI**: `r report_params$PI`

**Institution**: `r report_params$Institution`

**Department**: `r report_params$Department`

**Study Contact**: `r report_params$Study_Contact`

**Project Title**: `r report_params$Project_Title`

**Study Summary**: `r report_params$Study_Summary`

**Sample type(s)**: `r report_params$Sample_Types`

**Organism**: `r report_params$Organism`

**Analysis goal(s)**: `r report_params$Analysis_Goals`

**Report-prepared-by**:  
  - `r report_params$Report_Prepared_By`
  
**Report-reviewed-by**:  
  - `r report_params$Report_Reviewed_By`

# Data Downloads

## Download Raw Sequencing Data  

Below is a link to download the raw sequencing files. These files are very large (>150GB); download only when needed. Note that you **must be logged into your UF dropbox account** for this link to work.

```{r download raw data, echo=FALSE, results="asis"}
url <- report_params$raw_seq_URL

if (url == "") {
  cat("<p><em>This file was not included in this report.</em></p>")
} else {
  cat(sprintf('<a href="%s" class="btn btn-primary" role="button" download>Download Raw Sequence Files</a>', url))
}
```

## Download Sequencing Data Quality Control Summary

FastQC provides quality metrics for individual sequencing samples, including read quality, GC content, and adapter contamination. The MultiQC report summarizes these results across all samples. The multiQC report is of raw sequence data. These data were quality filtered and trimmed prior to analysis. For guidance on interpreting the results, see [FastQC documentation](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/).

```{r download multiqc, echo=FALSE, results="asis"}
url <- report_params$multiqc_url

if (url == "") {
  cat("<p><em>This file was not included in this report.</em></p>")
} else {
  cat(sprintf('<a href="%s" class="btn btn-primary" role="button" download>Download MultiQC Report</a>', url))
}
```

## Download Raw and Filtered Counts Data

Description of Data Sheets:

1. **Raw**: Raw gene count matrix before filtering or normalization.
2. **Filtered**: Gene expression counts filtered to remove lowly-expressed and non-variable genes using the `NOISeq` method. **This dataset was used as input for DE analysis.**
3. **CPM / CPM.batchcorrected**: Counts per million (CPM) normalized expression values, with gene symbols. If batch correction was applied (see methods), this sheet contains values back-transformed from the batch-corrected log-CPM matrix.
4. **LCPM / LCPM.batchcorrected**: Log-transformed CPM values, either uncorrected or batch corrected (if correction was requested), useful for visualization.
5. **Sample_Metadata**: Sample information including treatment, group, and library sizes.

The Excel file containing these datasets can be downloaded using the link provided below.

```{r download count data}
# Load necessary libraries
library(openxlsx)
library(dplyr)

# Extract count matrices from DGELists and add gene symbols
extract_dge_data <- function(dge) {
  df <- as.data.frame(dge$counts)
  df$ENSEMBL <- rownames(dge$counts)
  df <- left_join(df, dge$genes, by = "ENSEMBL")  # Add gene symbols
  df <- df %>% select(SYMBOL, everything())  # Move gene names to first column
  return(df)
}

dge_filtered_df <- extract_dge_data(dge_list_edgeR)
dge_raw_df <- extract_dge_data(dge_list_raw)
dge_NOISeq_df <- extract_dge_data(dge_list_NOISeq)

# Convert matrices to dataframes and add gene symbols
convert_matrix <- function(matrix, genes_df) {
  df <- as.data.frame(matrix)
  df$ENSEMBL <- rownames(matrix)
  df <- left_join(df, genes_df, by = "ENSEMBL")  # Add gene symbols
  df <- df %>% select(SYMBOL, everything())  # Move gene names to first column
  return(df)
}

# Determine whether to use batch-corrected log-CPM matrix
use_batch_corrected <- !is.null(report_params$batch_var) &&
                       report_params$batch_var == "Batch" &&
                       !is.null(dge_list_NOISeq$E_corrected)

if (use_batch_corrected) {
  message("Using batch-corrected matrix for both CPM and LCPM output.")
  lcpm_matrix <- dge_list_NOISeq$E_corrected
  cpm_matrix <- 2^lcpm_matrix  # Back-transform

  cpm_df <- convert_matrix(cpm_matrix, dge_list_edgeR$genes)
  lcpm_df <- convert_matrix(lcpm_matrix, dge_list_edgeR$genes)

  cpm_name <- "CPM.batchcorrected"
  lcpm_name <- "LCPM.batchcorrected"
} else {
  lcpm_matrix <- cpm(dge_list_NOISeq, log = TRUE)
  cpm_matrix <- cpm(dge_list_NOISeq)

  cpm_df <- convert_matrix(cpm_matrix, dge_list_edgeR$genes)
  lcpm_df <- convert_matrix(lcpm_matrix, dge_list_edgeR$genes)

  cpm_name <- "CPM"
  lcpm_name <- "LCPM"
}

# Extract sample metadata
sample_metadata <- dge_list_NOISeq$samples

# Create a list of dataframes to be written to an Excel file
df_list <- list(
  "Raw" = dge_raw_df,
  "Filtered" = dge_NOISeq_df,
  cpm_df,
  lcpm_df,
  "Sample_Metadata" = sample_metadata
)

# Rename CPM and LCPM elements after the list is created
names(df_list)[3:4] <- c(cpm_name, lcpm_name)

if (!is.null(report_params$batch_var) && report_params$batch_var == "Batch" && !is.null(dge_list_NOISeq$E_corrected)) {
  cat("The CPM and LCPM matrices have been batch-corrected using limma::removeBatchEffect. These corrected matrices are labeled CPM.batchcorrected and LCPM.batchcorrected in the downloadable Excel file.")
} else {
  cat("Note:No batch correction was applied. CPM and LCPM matrices are based on the standard log-transformed CPM from filtered counts.")
}

# Generate a download button
df_list %>% downloadthis::download_this(
  output_name = "RNAseq_Data_Analysis",
  output_extension = ".xlsx",
  button_label = "Download RNA-seq Data File",
  button_type = "primary",
  has_icon = TRUE,
  icon = "fa fa-save"
)
```

# Data Visualizations

## PCA (multivariate analaysis)</span>

Principal component analysis of gene expression data were calculated in R v4.3.3. Two-dimensional PCA score plots reveal possible separation in gene expression profiles and can help to identify quality control issues and batch effects. Ellipses are calculated using the R package car (Fox J. and Weisberg S. 2019) and ~1 Std dev.

### PCA of all samples

If batch correction was applied using limma::removeBatchEffect(), the plots below compare PCA of the filtered expression matrix before and after correction. This allows visualization of how batch effects were reduced while preserving biological signal.

```{r pca all, fig.width=12, fig.height=4}
genes_filtered <- dge_list_NOISeq$genes[rownames(dge_list_NOISeq$counts), , drop = FALSE]
if (!is.null(report_params$batch_var) && report_params$batch_var == "Batch") {
  dge_uncorrected <- edgeR::DGEList(
    counts = dge_list_NOISeq$counts,
    samples = dge_list_NOISeq$samples,
    genes = genes_filtered
  )

  dge_corrected <- edgeR::DGEList(
    counts = 2^dge_list_NOISeq$E_corrected,
    samples = dge_list_NOISeq$samples,
    genes = genes_filtered
  )

  dge_list <- list(dge_uncorrected, dge_corrected)
  annotation_labels <- c("Filtered (Uncorrected)", "Filtered (Batch Corrected)")
} else {
  dge_list <- list(dge_list_raw, dge_list_NOISeq, dge_list_edgeR)
  annotation_labels <- c("Raw", "Filtered (NOISeq)", "Filtered (edgeR)")
}


plot_pca_combined(dge_list, grp_var = report_params$group_var, annotation_labels = annotation_labels)
if (!is.null(report_params$batch_var) && report_params$batch_var == "Batch") {
plot_pca_combined(dge_list, grp_var = "Batch", annotation_labels = annotation_labels)}
```

### PCA per-contrast

```{r pca,eval=FALSE}
# Apply PCA separately for each contrast
pca_results <- lapply(names(efit_results_dfs)[1:2], function(contrast_name) {
  plot_pca_by_contrast(dge_list = list(dge_list_NOISeq), 
                        contrast_name = gsub("_results_df","",contrast_name), 
                        group_var = report_params$group_var)
})

# Display PCA plots
tagList(pca_results)
```

```{r contrast pcas, fig.width=8, fig.height=8, message=FALSE, warning=FALSE}
# Generate list of PCA plots (ggplot objects)
pca_results <- lapply(names(efit_results_dfs), function(contrast_name) {
  plot_one_pca_by_contrast(
    dge_list = dge_list_NOISeq,
    contrast_name = gsub("_results_df", "", contrast_name),
    group_var = report_params$group_var
  )
})

# Display PCA plots
tagList(pca_results)
```

## Global comparison of transcriptional profile similarity between groups

### Spearman correlation

Pairwise sample-to-sample correlation coefficients of log CPM normalized filtered counts were calculated using Spearman’s rank correlation in R v4.3.3 and visualized using pheatmap 1.0 (Kolde 2019).

```{r corr,fig.width=12, fig.height=12}
library(pheatmap)
library(tibble)
library(RColorBrewer)
library(grid)

# Filter samples
if (!is.null(report_params$filter_samples) && report_params$filter_samples != ""){
  selected_samples <- sample.keys %>%
    filter(!SampleName %in% report_params$filter_samples) %>%
    pull(SampleName)
} else {
  selected_samples <- sample.keys %>% pull(SampleName)
}

# Subset lcpm
lcpm_filtered <- lcpm_matrix[, colnames(lcpm_matrix) %in% selected_samples]

# Spearman correlation
cor_matrix <- cor(lcpm_filtered, method = "spearman")

# Annotations
annotation_col <- sample.keys %>%
  filter(SampleName %in% colnames(cor_matrix)) %>%
  select(SampleName, all_of(report_params$group_var)) %>%
  column_to_rownames("SampleName")

# Dynamic color palette
group_levels <- sort(unique(annotation_col[[report_params$group_var]]))
n_groups <- length(group_levels)

if (n_groups <= 8) {
  color_palette <- brewer.pal(n_groups, "Set2")
} else {
  color_palette <- colorRampPalette(brewer.pal(8, "Set2"))(n_groups)
}
ann_colors <- setNames(color_palette, group_levels)

# Correlation color scale
red_palette <- colorRampPalette(c("blue", "white", "red"))(100)
min_val <- min(cor_matrix)
max_val <- max(cor_matrix)
breaks_seq <- seq(min_val^2, max_val^2, length.out = 101)^(1/2)

# Automatically adjust plot size
n_samples <- ncol(cor_matrix)
display_numbers_flag <- n_samples <= 40
fig_height <- max(6, n_samples * 0.25)
fig_width <- max(6, n_samples * 0.25)

# Draw with layout control
p <- pheatmap(
  cor_matrix,
  color = red_palette,
  breaks = breaks_seq,
  annotation_col = annotation_col,
  annotation_row = annotation_col,
  annotation_colors = setNames(list(ann_colors), report_params$group_var),
  clustering_method = "complete",
  main = paste("Spearman's Correlation Plot for Normalized Read Counts (lcpm) filtered by", report_params$group_var),
  display_numbers = display_numbers_flag,
  number_format = "%.3f",
  fontsize = 10,
  fontsize_number = 6,
  number_color = "black",
  fontsize_row = 6,
  fontsize_col = 6,
  silent = TRUE  # return the object instead of drawing immediately
)

# Force proper rendering in Rmd
grid.newpage()
grid.draw(p$gtable)

```

### Heatmap of top 1000 CV genes

Scaled gene expression values of the top 1000 variable genes (sd/mean) were visualized using pheatmap 1.0 (Kolde 2019) with sample and gene clustered by complete hierarchical clustering.

```{r,fig.width=12,fig.height=6}
# Step 3: Identify the top 1000 most variable genes by CV
# Step: Identify top 1000 most variable genes by CV
cv_values <- apply(lcpm_filtered, 1, function(x) sd(x) / mean(x))
top_1000_genes <- names(sort(cv_values, decreasing = TRUE))[1:1000]
lcpm_top1000 <- lcpm_filtered[top_1000_genes, ]

# Step: Plot
pheatmap(
  lcpm_top1000,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  annotation_col = annotation_col,
  annotation_colors = setNames(list(ann_colors), report_params$group_var),
  clustering_method = "complete",
  scale = "row",
  show_rownames = FALSE,
  show_colnames = TRUE,
  main = "Top 1000 Most Variable Genes' (CV) Expression (Z-score) by Sample",
  fontsize = 10
)
```

## Mean-Variance Relationship: Before and After Variance Moderation

These **mean-variance trend** plots are provided to illustrate how the DE analysis ensure differential expression (DE) results are **not biased by expression-dependent variance**. The top plot shows how **variance changes with expression level** in the raw data. Typically, variance **decreases** as expression increases. This confirms that `voom()` correctly models **heteroscedasticity (unequal variance)** in RNA-seq data. The **Bottom plot** shows variance **after empirical Bayes shrinkage** using `eBayes()`. The trend should be **flat**, indicating variance is properly stabilized. 

```{r SA plot side by side, fig.width=8, fig.height=4, out.width="100%"}
library(gridExtra)
library(grid)
library(png)

# Load the PNG plot as a raster image
img <- rasterGrob(readPNG(voom_plot_path), interpolate = TRUE)
# Open a graphics device, plot `plotSA()`, then capture it
tmp <- tempfile(fileext = ".png")
invisible(png(tmp, width = 600, height = 600, res = 150))  # Open a temporary PNG device
par(mar=c(3, 5, 2, 2), cex.axis=0.7, cex.lab=0.7, cex.main=0.7)  
plotSA(efit_list[[1]], main = "Final model: Mean-Variance trend")
invisible(dev.off())  # Close the device

# Load the saved plot as a raster image
plot_sa_grob <- rasterGrob(readPNG(tmp), interpolate = TRUE)

# Arrange both images side by side
grid.arrange(img, plot_sa_grob, ncol = 2)
```

# Differential Expression Analysis

Differential expression analysis was performed using limma v3.5 (Ritchie et al. 2015). Raw RNA-seq counts were filtered using NOISeq v2.4 (Tarazona et al., 2011) to remove features that have an average expression per condition less than 1 cpm a coefficient of variation > 100% in all the conditions. To account for differences in sequencing depth and composition bias, TMM normalization as implemented in edgeR v3.4 (Robinson et al. 2010) was used to calculate normalization factors that were used during voom transformation. A linear model was fitted to the transformed data, and pairwise comparisons between conditions were assessed by computing moderated t-statistics, moderated F-statistic, and log-odds of differential expression with empirical Bayes moderation of the standard errors. Statistical significance was assessed using adjusted p-values to control the false discovery rate (FDR).

```{r,results='asis'}
# Print the number of contrasts tested
cat(paste0(length(names(efit_list)), " contrasts were tested:"))

# Create a table of contrasts
contrast_table <- data.frame("Contrast" = names(efit_list))
kable(contrast_table, format = "html", caption = "Table: Tested Contrasts")
```

## Summary of DE results per contrast

Down-regulated: adj. p value < 0.05 **AND** logFC > 0.58

Up-regulated: adj. p value < 0.05 **AND** logFC < -0.58

Not Significant: adj. p value > 0.05 **AND/OR** abs(logFC) > 0.58

```{r DE_summary_all, echo=FALSE, results='asis'}
tagList(
  lapply(seq_along(efit_list), function(i) {
    display_de_summary(efit_list[[i]],contrast=names(efit_list)[i])  # Directly return the datatable
  })
)
```

## Interactive Downloadable Differential Expression Results Tables

- Summary per contrast showing the number of differentially expressed genes and key statistics.
- **Tables can be filtered, sorted, and downloaded**  

```{r download tables}
# Convert efit_list to a named list of data frames with contrast column
efit_df_list <- lapply(seq_along(efit_list), function(i) {
  contrast_name <- gsub("^X| X", "", names(efit_list)[i])
  df <- download_de_result_table(efit_list[[i]])
  df$contrast <- contrast_name
  return(df)
})

names(efit_df_list) <- gsub("^X| X", "", names(efit_list))

# Download GO results as an Excel file
efit_df_list %>% downloadthis::download_this(
  output_name = paste0(report_params$Project_Title,"DE_results_all_contrasts"),
  output_extension = ".xlsx",
  button_label = "Download DE results for all contrasts",
button_type = "primary",
  has_icon = TRUE,
  icon = "fa fa-save"
)
```

```{r view tables}
tagList(
  lapply(seq_along(efit_list), function(i) {
    contrast_name <- gsub("^X| X", "", names(efit_list)[i])
    display_de_result_table(efit_list[[i]], contrast_name = contrast_name)  # Directly return the datatable
  })
)
```

## Differential Expression Results Visualizations

### Volcano Plots

Visualization of **log-fold changes vs. statistical significance** for each contrast (plotly v4.1 (Sievert C 2020)) shows patterns of **significantly up/downregulated genes** ("significance" determined by adj. p value < 0.05 ***AND** abs(logFC) > 0.58).

```{r volcano_plots_all, echo=FALSE, results='asis', fig.width=9, fig.height=5}
# Generate all volcano plots and store in a list
volcano_plots <- lapply(seq_along(efit_results_dfs), function(i) {
  generate_volcano_plot(efit_results_dfs[[i]], paste0(gsub("efit_|_results_df","",names(efit_results_dfs)[i]), "\n")) %>%
    style(marker = list(size = 2))
})

# Render all plots
tagList(volcano_plots)
```

### Heatmaps

Heatmaps of the **top 50 DE genes** for each contrast show the scaled counts values for all samples in that contrast using heatmaply v1.5 (Galili et al. 2017). Hierarchical clustering of samples is based on expression patterns.

```{r heatmaps of top DE genes,fig.width=9, fig.height=9}
heatmap_list <- lapply(seq_along(efit_results_dfs), function(i) {
  generate_heatmap(
    efit_results_df = efit_results_dfs[[i]],  
    lcpm_matrix = lcpm_matrix,  
    dge_list_NOISeq = dge_list_NOISeq,  
    title = paste("Scaled Gene Expression (Z-scores of logCPM) of Top 50 DE Genes\n", gsub("efit_|_results_df","",names(efit_results_dfs)[i])),  
    num_genes = 50,  
    fontsize_row = 9
  )
})
tagList(heatmap_list)
```

# Pathway Enrichment Analysis

Pathway enrichment analysis was performed to identify functional enrichment of gene lists and to compare these significant results across contrasts.

If a contrast had fewer than five significant genes using the standard filtering (p adj < 0.05; abs(log2FC) > 0.58), relaxed filtering was applied to the DE results (adj.P < 0.1, No log2FC filtering) to create an input gene list for the enrichment analyses. If no genes passed standard or relaxed filtering criteria, the enrichment analysis was not performed for that contrast-direction (see warning messages).

```{r gene-list-and-messages, eval=TRUE}
# Generate dynamic gene lists and collect enrichment messages
enrichment_messages <- c()

for (i in seq_along(entrez_ids_list)) {
  contrast_name <- names(entrez_ids_list)[i] %>%
    gsub("top_DE_entrezIDs_?", "", .)

  up_entry <- entrez_ids_list[[i]]$up
  down_entry <- entrez_ids_list[[i]]$down

  if (!is.null(up_entry$message)) enrichment_messages <- c(enrichment_messages, paste0(gsub("efit_|_results_df","",contrast_name), " (up): ", up_entry$message))
}
```

```{r enrichment-messages, results='asis', echo=FALSE}
if (length(enrichment_messages) > 0) {
  cat("### ⚠️ Notes on DE gene thresholds\n\n")
  cat(paste0(
    "<ul>",
    paste0("<li>", enrichment_messages, "</li>", collapse = ""),
    "</ul>"
  ))
}
```

## Gene Ontology (GO) Enrichment Analysis

Gene Ontology enrichment was performed using the enrichGO function in clusterProfiler v4.8 (Yu et al. 2012) in each of three GO categories (BP, MF, CC). **Interactive GO enrichment plots** are based on the top 10 significantly enriched GO terms per GO category per gene list. Hover over the plot to view p-value, gene ratio, and up to the top 20 DE genes (sorted by DE adj.p.value) in that term.

To assess similarities between gene lists' enrichment results, if a gene list(s) had significant results for a different gene lists' top-10 term, that result is displayed as well regardless of whether or not the result was in top 10. **Downloadable results excel file** include all significant results for all gene lists.

### DE genes

Gene lists are significantly DE genes per contrast

```{r GO enrichment,eval=TRUE}
# Generate dynamic gene lists
gene_lists <- list()

for (i in seq_along(entrez_ids_list)) {
  contrast_name <- names(entrez_ids_list)[i] %>%
    gsub("top_DE_entrezIDs_?", "", .)

  up_entry <- entrez_ids_list[[i]]$up
  down_entry <- entrez_ids_list[[i]]$down

  # Add upregulated genes
  if (length(up_entry$entrez_ids) > 0) {
    gene_lists[[paste0(contrast_name, ".up")]] <- up_entry$entrez_ids
  }

  # Add downregulated genes
  if (length(down_entry$entrez_ids) > 0) {
    gene_lists[[paste0(contrast_name, ".down")]] <- down_entry$entrez_ids
  }
}
```

```{r run GO}
# Perform GO Enrichment for BP, MF, CC
GO_BP_results <- generate_enrichment_plot(
  gene_lists = gene_lists, 
  de_results_df = bind_rows(efit_results_dfs, .id = "contrast"), 
  universe_entrez = universe_entrez,
  ont_category = "BP",
  annotation_db = report_params$annotation_db
)

GO_MF_results <- generate_enrichment_plot(
  gene_lists = gene_lists, 
  de_results_df = bind_rows(efit_results_dfs, .id = "contrast"), 
  universe_entrez = universe_entrez,
  ont_category = "MF",
  annotation_db = report_params$annotation_db
)

GO_CC_results <- generate_enrichment_plot(
  gene_lists = gene_lists, 
  de_results_df = bind_rows(efit_results_dfs, .id = "contrast"), 
  universe_entrez = universe_entrez,
  ont_category = "CC",
  annotation_db = report_params$annotation_db
)
```

```{r GO enrich plots,fig.width=12, fig.height=10}
# Display interactive plots
download_button_png(GO_BP_results$static_plot, "GO_BP_enrich_plot",height = 12)
GO_BP_results$interactive_plot

download_button_png(GO_MF_results$static_plot, "GO_MF_enrich_plot",height=12)
GO_MF_results$interactive_plot

download_button_png(GO_CC_results$static_plot, "GO_CC_enrich_plot",height=12)
GO_CC_results$interactive_plot

# Combine GO results into a single list of data frames (some might be NULL)
GO_results_list <- list(
  "BP_Results" = GO_BP_results$go_results,
  "MF_Results" = GO_MF_results$go_results,
  "CC_Results" = GO_CC_results$go_results
)

# Remove NULL elements
GO_results_list <- GO_results_list[!sapply(GO_results_list, is.null)]

# Download GO results as an Excel file if any exist
if (length(GO_results_list) > 0) {
  GO_results_list %>% downloadthis::download_this(
    output_name = "GO_results",
    output_extension = ".xlsx",
    button_label = "Download GO enrichment results for all GO categories and all contrasts",
    button_type = "primary",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
} else {
  cat("<b>No GO enrichment results available to download.</b>")
}
```

## KEGG Pathway Enrichment Analysis

KEGG enrichment was performed using the enrichKEGG function in clusterProfiler v4.8 (Yu et al. 2012). **Interactive KEGG enrichment plots** are based on the top 10 significantly enriched KEGG pathways per gene list. Hover over the plot to view p-value, gene ratio, and up to the top 20 DE genes (sorted by DE adj.p.value) in that pathway.

To assess similarities between gene lists' enrichment results, if a gene list(s) had significant results for a different gene lists' top-10 pathway, that result is displayed as well regardless of whether or not the result was in top 10. **Downloadable results excel file** include all significant results for all gene lists.

```{r KEGG,eval=TRUE}
# Run the function
kegg_results <- generate_kegg_enrichment_plot(
  gene_lists = gene_lists, 
  de_results_df = bind_rows(efit_results_dfs, .id = "contrast"), 
  universe_entrez = universe_entrez,
  annotation_db = report_params$annotation_db 
)
```

### DE genes

Gene lists are significantly DE genes per contrast

```{r kegg plot,fig.width=12, fig.height=10}
download_button_png(kegg_results$static_plot, "kegg_enrichment_plot",height=12)
kegg_results$interactive_plot


kegg_results$kegg_results %>%  downloadthis::download_this(
  output_name = "KEGG_results",
  output_extension = ".xlsx",
  button_label = "Download KEGG enrichment results for all contrasts",
button_type = "primary",
  has_icon = TRUE,
  icon = "fa fa-save"
)
```

```{r cleanup_temp_dir, echo=FALSE, message=FALSE}
if (!is.null(params$out_dir) && dir.exists(params$out_dir)) {
  unlink(params$out_dir, recursive = TRUE)
}
```


# References

- Robinson MD, McCarthy DJ and Smyth GK (2010). edgeR: a Bioconductor package for differential expression analysis of digital gene expression data. Bioinformatics
  26, 139-140
- Tarazona S, García-Alcalde F, Dopazo J, Ferrer A, Conesa A (2011). *Differential expression in RNA-seq: a matter of depth*. Genome Research, 21(12), 2213-2223.  
- Ritchie ME, Phipson B, Wu D, Hu Y, Law CW, Shi W, Smyth GK (2015). “limma powers differential expression analyses for RNA-sequencing and microarray studies.” Nucleic Acids Research, 43(7), e47. doi:10.1093/nar/gkv007.
- Yu G, Wang L, Han Y, He Q (2012). “clusterProfiler: an R package for comparing biological themes among gene clusters.” OMICS: A Journal of Integrative Biology, 16(5), 284-287. doi:10.1089/omi.2011.0118.
- R Core Team (2021). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria.
- Kolde R (2019). pheatmap: Pretty Heatmaps. R package version 1.0.12, <https://CRAN.R-project.org/package=pheatmap>.
- Galili, Tal, O'Callaghan, Alan, Sidi, Jonathan, Sievert, Carson (2017). “heatmaply: an R package for creating interactive cluster heatmaps for online publishing.” Bioinformatics. 
- Sievert C (2020). Interactive Web-Based Data Visualization with R, plotly, and shiny. Chapman and Hall/CRC. ISBN 9781138331457, https://plotly-r.com.



